LWE Trapdoor One-Way Function

### 2017-07-20 latex/qtpie/paper.tex
more paper work

\begin{abstract}
 We define a simple trapdoor function based on the Learning-With-Errors problem in one dimension, and present a key exchange algorithm based on the trapdoor. The trapdoor algorithm is efficient in terms of key size, key generation time, ciphertext size, and public/private key operation time: We can perform many thousands of public and private key operations per second with a naive python implementation on modest hardware. 
 \end{abstract}

@@

In this paper we define a candidate TOWF based on 1-dimensional LWE. It is based on a simple observation of a congruence, $a s + e \equiv s + a^{-1} e \mod P$, along with some subtle manipulation of the sizes of the numbers involved. The trapdoor information is the short inverse $a^{-1}$, which allows the recovery of $s, e$ from a point $a s + e \mod P$, provided some constraints on the sizes of numbers are satisfied.

We use the TWOF to construct basic public key algorthims for exchanging keys and creating digital signatures.

\subsection{Definitions}
We $+$ to denote addition, $a b$ to indicate multiplication, $a / b$ to denote division, and $\mod p$ to denote the modulo operation. We use $\leftarrow$ to indicate the assignment of a value to a name, and $\rightarrow$ to indicate the sending of data from one party to another. We use the function $inverse(k, n)$ to indicate the computation of the modular multiplicative inverse of $k \mod n$. We use the function $random(k)$ to indicate the generation of of a random $k$-bit integer. We write application of the public key operation as $key_{pub}(s,\ e)$, and application of the private key operation as $key_{priv}(c)$.

\section{Algorithm}
There is a publicly agreed upon modulus $P$. All math is done modulo $P$. The same $P$ may always be used. It should be prime; It it is not secret, and otherwise only needs to be both agreed upon and large enough. The size of $P$ is determined by the rest of the parameter sizes which are discussed below; In our python implementation which targets a 256-bit security level, $P$ is slightly less then 800 bits in size. The larger $P$ is, the longer the BigNum arithmetic operations take, and the larger ciphertexts become; Consequently, we opt for the smallest $P$ possible.

@@

Given $c, p, N$, we can recover $s, e$ using the following algorithm:

@@

\begin{flalign*}
key_{pub} \leftarrow a r\\
s \leftarrow random(s_{size})\\
e \leftarrow random(e_{size})\\
c \leftarrow key_{pub} s + e\\
key_{pub}^{-1} \leftarrow inverse(key_{pub}, P)\\
c_i \leftarrow c key_{pub}^{-1}\\
c_i \equiv s + a^{-1} r^{-1} e\\
\end{flalign*}

In the final line $c_i \equiv s + a^{-1} r^{-1} e$, we can see that the inverse of the public key is too large if $a^{-1} r^{-1} e > P$; In such a case, $s + a^{-1} r^{-1} e \mod a^{-1} r^{-1}$ will not yield $s$; While $a^{-1}$ is relatively small, $r^{-1}$ is (most likely) large - Ultimately, the size of $a^{-1} r^{-1}$ is roughly the same as $a r$, and so this alternative perspective of $s + a^{-1} r^{-1} e$ offers no more advantage to recovering $s, e$ as then the original $ars + e$ did.

@@

\begin{flalign*}
s \leftarrow random(s_{size})\\
e \leftarrow random(e_{size})\\
c \leftarrow key_{pub}s + e\\
c_i \leftarrow c a^{-1}\\
c_i \equiv sr + a^{-1} e
\end{flalign*}

@@

\subsection{Key Generation}
\begin{flalign*}
key_{priv} \leftarrow random(priv_{size})\\
a \leftarrow inverse(key_{priv}, N)\\
r \leftarrow random(r_{size})\\
k_{pub} \leftarrow a r\\
\end{flalign*}

@@

\subsection{Public Key Operation}
\begin{flalign*}
a \leftarrow key_{pub}\\
s \leftarrow random(s_{size})\\
e \leftarrow random(e_{size})\\
c \leftarrow a s + e\\
\end{flalign*}

@@

\section{Key Exchange and Digital Signatures}
\subsection{Key Exchange}
Given a TWOF, a basic algorithm to exchange keys can be constructed as follows: Assuming that Alice has Bob's public key already, she simply applies the public key operation to a random value and sends the result to Bob; Bob can then use the private key to recover the random value. Assuming that both parties are honest, then the random value will be unknown to all parties but Alice and Bob, and they can use this information to construct a secured communications channel.

@@

@@

We note that ciphertexts produced by the TWOF are additively homomorphic - this means that signatures will possess this quality as well. This feature can be removed by signing a hash of the message instead of the message itself.

@@

\subsection{Key Generation}
It would appear that obtaining the private key from the public key must be done by guessing one of $a, a^{-1}, r$. If this were not the case, then whatever algorithm that can be used to recover any of $a, a^{-1}, r$ from $a r \mod P$ could be used to recover plaintexts from ciphertexts of the form $m k \mod P$ for random $m, k$ where one of $m, k$ is smaller then the other (but still large enough to reduce modulo $P$). It seems unlikely that such an attack exists and is faster then guessing $a, a^{-1},  r$.

@@

If we consider the case where $ar = 1$, then $ars + e \equiv s + e \mod P$ for uniformly random values $s, e$.

@@

\section{Metrics}
Using our public domain python implementation that is configured for a 256-bit security level, we can perform 10,000 public key operations in slightly under 1 second. We can perform roughly 30,000 private key operations per second. Our tests were performed on modest hardware, namely an Intel Celeron N2830 2.16GHz with 2 cores.

@@

\subsection{Implementation Complexity}
The implementation complexity of the scheme is very low. Our public domain python implementation of the algorithm spans barely 100 lines, including the documentation and comments.\\

Additional code is required for supporting functions such as the generation of prime numbers (only if $N$ is being generated/not already established) and unit testing. These are modest requirements that can be implemented without much complexity. \\

\section{Conclusion}
We define a conceptually simple, resource-practical Trapdoor One-Way Function. 
\end{document}

>>>>

\begin{abstract}
 We define a simple trapdoor function based on the Learning-With-Errors problem in one dimension. The trapdoor algorithm is time and space efficient: We can perform tens of thousands of public and private key operations per second with a naive python implementation on modest hardware, and the keys and ciphertexts are small.
 \end{abstract}

@@

In this paper we define a candidate TOWF based on 1-dimensional LWE. It is based on a simple observation of a congruence, $a^{-1}(a s + e) \equiv s + a^{-1} e \mod P$, along with some subtle manipulation of the sizes of the numbers involved. The trapdoor information is the short inverse $a^{-1}$, which allows the recovery of $s, e$ from a point $a s + e \mod P$, provided some constraints on the sizes of numbers are satisfied.

We discuss the possibilities to use the TWOF to construct basic public key algorthims for encapsulating keys, public key encryption, and creating digital signatures. We go on to provide statistics for a key encapsulation mechanism, gathered using our open source public domain implementation of the algorithm. We conclude with some open problems about formal security proof for the algorithm.

\subsection{Definitions}
We $+$ to denote addition, $a b$ to indicate multiplication, $a / b$ to denote division, and $\mod p$ to denote the modulo operation. We use $\leftarrow$ to indicate the assignment of a value to a name. We use the function $inverse(k, n)$ to indicate the computation of the modular multiplicative inverse of $k \mod n$. We use the function $random(k)$ to indicate the generation of of a random $k$-bit integer. We write application of the public key operation as $key_{pub}(s,\ e)$, and application of the private key operation as $key_{priv}(c)$.

\section{Algorithm}
There is a publicly agreed upon modulus $P$. All math is done in the field of integers modulo $P$. The same $P$ may always be used. It should be prime; It it is not secret, and otherwise only needs to be both agreed upon and large enough. The size of $P$ is determined by the rest of the parameter sizes which are discussed below; In our python implementation which targets a 256-bit security level, $P$ is about 783 bits in size. The larger $P$ is, the longer the BigNum arithmetic operations take, and the larger ciphertexts become; Consequently, we opt for the smallest $P$ possible.

@@

We refer to this expression as the public key operation. If the expression $a s + e$ were over integers instead of modulo $P$, then we could recover $e$ simply by performing $c \mod a$; We could similarly recover $s$ by performing $c / a$. However, since $a s > P$ then this will no longer work.

Given $c, a, P$, we can recover $s, e$ using the following algorithm:

@@

\begin{flalign*}
k_{pub} \leftarrow a r\\
s \leftarrow random(s_{size})\\
e \leftarrow random(e_{size})\\
c \leftarrow k_{pub} s + e\\
k_{pub}^{-1} \leftarrow inverse(k_{pub}, P)\\
c_i \leftarrow k_{pub}^{-1} c\\
c_i \equiv s + a^{-1} r^{-1} e\\
\end{flalign*}

In the final line $c_i \equiv s + a^{-1} r^{-1} e$, we can see that the inverse of the public key will be too large if $a^{-1} r^{-1} e > P$; In such a case, $s + a^{-1} r^{-1} e \mod a^{-1} r^{-1}$ will not yield $s$; While $a^{-1}$ is relatively small, $r^{-1}$ is (most likely) large - Ultimately, the size of $a^{-1} r^{-1}$ is roughly the same as $a r$, and so this alternative perspective of $s + a^{-1} r^{-1} e$ offers no more advantage to recovering $s, e$ as then the original $ars + e$ did.

@@

\begin{flalign*}
\dots\\
c_i \leftarrow c a^{-1}\\
c_i \equiv sr + a^{-1} e
\end{flalign*}

@@

\subsection{Key Generation}
\begin{flalign*}
k_{priv} \leftarrow random(priv_{size})\\
a \leftarrow inverse(k_{priv}, N)\\
r \leftarrow random(r_{size})\\
k_{pub} \leftarrow a r\\
\end{flalign*}

@@

\subsection{Public Key Operation}
\begin{flalign*}
a \leftarrow k_{pub}\\
s \leftarrow random(s_{size})\\
e \leftarrow random(e_{size})\\
c \leftarrow a s + e\\
\end{flalign*}

@@

\section{Key Encapsulation, Public Key Encryption, and Digital Signatures}
\subsection{Key Encapsulation}
Given a TWOF, a basic algorithm to exchange keys can be constructed as follows: Assuming that Alice has Bob's public key already, she simply applies the public key operation to a random value and sends the result to Bob; Bob can then use the private key to recover the random value. Assuming that both parties are honest, then the random value will be unknown to all parties but Alice and Bob, and they can use this information to construct a secured communications channel.

@@

\subsection{Public Key Encryption}
We can generalize and say that the Key Encapsulation Mechanism presented in the prior section is an example of using the trapdoor for public key encryption on random plaintext values. It is clear then that one of the values of $s, e$ can be used to store a plaintext value.

Considering the efficiency of hybrid cryptosystems, where an asymmetric algorithm is used to establish a key for a message, then a symmetric cipher is used to perform the actual encryption, we first must determine what can be gained by storing a message inside the public key ciphertext.

The most attractive feature is probably the homomorphism - the ability to add and manipulate ciphertexts can enable more complex cryptographic schemes and protocols. However, the manner with which plaintext information is stored and possibly padded will influence the nature of the homomorphic properties, and a careful configuration will be required to preserve the desired relationship(s). For example, if $e$ is used to store the message, then plaintext adjustments may be made to the resultant ciphertext; However, if $s$ is used as to store the message, then the ciphertext will no longer facilitate plaintext adjustments.

@@

We note that ciphertexts produced by the TWOF are additively homomorphic - this means that signatures will possess this quality as well. Usually this feature can be removed by incorporating a hash function into the signature process and padding the result before applying the private key signing operation.

@@

\subsection{Key Generation}
Obtaining the private key from the public key must be done by guessing one of $a, a^{-1}, r$. If this were not the case, then whatever algorithm that can be used to recover any of $a, a^{-1}, r$ from $a r \mod P$ could be used to recover plaintexts from ciphertexts of the form $m k \mod P$ for random $m, k$ where one of $m, k$ is smaller then the other (but still large enough to reduce modulo $P$). It seems unlikely that such an attack exists and is faster then guessing $a, a^{-1},  r$.

@@

We will first consider the case where $e \leftarrow random(1)$. The task of the adversary is to recover $e$, given $a,\ a s + e$ for $a \leftarrow \{0,\ \dots,\ P\}$, $s \leftarrow \{0,\ \dots,\ P\}$. If an adversary can guess the least significant bit of $a s$, then they can obtain $e$ from $a s + e$.

If we consider the case where $a = 1$, then $a s + e \equiv s + e \mod P$ for uniformly random values $s, e$. Recovering $s,\ e$ in this situation is clearly intractable.

If we consider 8-bit $a, s$ values and $P = 257$, then we can evaluate the function $a x$ on values $\{0,\dots,\ 256\}$ and produce an 8-bit s-box, like the kind used in symmetric cryptography. We can then analyze the result with the same techniques and tools. We have done so and found differentials ranging from probability $62/256$ in the worst case, to around $64/256$ in the average case, to $254/256$ in the worst case. However, considering that each ciphertext $a s + e$ uses a unique, random $s$, it is not clear how to select a differential to try to exploit. Additionally, real instances of the problem will use $a,\ s$ values that are much larger then any tool for analyzing s-boxes can handle.

@@

\section{Metrics}
Using our public domain python implementation that is configured for a 256-bit security level, we can perform 10,000 public key operations in 1 second. We can perform roughly 30,000 private key operations per second. Our tests were performed on modest hardware, namely an Intel Celeron N2830 2.16GHz with 2 cores.

@@

\section{Open Problems}
Our results are lacking in rigorous proof; We do not posses the requisite skills to create such proofs. We leave it as an open problem to formally prove (or dis-prove) the security of the key generation function and public key operation.

We do not know if the private key operation can be proven secure; We leave this as an open problem, along with the construction of a concrete digital signature scheme.

\section{Conclusion}
We define a conceptually simple, resource-practical Trapdoor One-Way Function. We discuss the potential applications of such a function, and provide performance statistics from our open-source public-domain python implementation of a key encapsulation mechanism using the trapdoor function. Finally, we discuss some open problems related to formally proving the security of the algorithm.
\end{document}
